---
vignette: >
  % \VignetteIndexEntry{BioNet maximum scoring subnetwork analysis}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup
```{r, message = FALSE}
library(BioPlex)
library(BioNet)
library(EnrichmentBrowser)
library(graph)
```

# BioNet

Get the latest version of the 293T PPI network:

```{r}
bp.293t <- BioPlex::getBioPlex(cell.line = "293T", version = "3.0")
```

and turn into a graph object:

```{r}
bp.gr <- BioPlex::bioplex2graph(bp.293t)
```

Get RNA-seq data for 293T cells from GEO: 
[GSE122425](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122425)

```{r gse122425}
gse122425 <- BioPlex::getGSE122425()
```

Differential expression analysis between NSUN2 knockout and wild type: 

```{r de-ana-gse122425}
gse122425$GROUP <- rep(c(0, 1), each = 3)
gse122425 <- EnrichmentBrowser::deAna(gse122425, de.method = "DESeq2")
```

Inspect results:

```{r de-results-gse122425}
rowData(gse122425)
not.na <- !is.na(rowData(gse122425)$PVAL)
gse122425 <- gse122425[not.na,]
EnrichmentBrowser::volcano(rowData(gse122425)$FC, rowData(gse122425)$ADJ.PVAL)
```

Change rownames to gene symbols for intersection with bioplex graph:
 
```{r}
rowData(gse122425)$ENSEMBL <- rownames(gse122425)
rownames(gse122425) <- rowData(gse122425)$SYMBOL
```

Intersect bioplex graph and expression dataset:

```{r isect-bp-gse122425}
syms <- graph::nodeData(bp.gr, graph::nodes(bp.gr), "SYMBOL")
syms <- unlist(syms)
isect <- intersect(syms, rownames(gse122425)) 
ind <- match(isect, syms) 
bp.sgr <- graph::subGraph(names(syms)[ind], bp.gr)
bp.sgr 
bp.sgr <- BioNet::rmSelfLoops(bp.sgr)
bp.sgr
```

Get the DE p-values for each node:

```{r de-bp-gse122425}
syms <- syms[ind]
ind <- match(isect, rownames(gse122425))
pvals <- rowData(gse122425)[ind, "PVAL"]
names(pvals) <- names(syms)
head(pvals)
```

To score each node of the network we fit a Beta-uniform mixture model (BUM)
to the p-value distribution and subsequently use the parameters of the model 
for the scoring function:

```{r bum-bp-gse122425}
fb <- BioNet::fitBumModel(pvals)
scores <- BioNet::scoreNodes(bp.sgr, fb, fdr = 0.05)
head(scores)
summary(scores)
```

## First reduction step: unsupervised detection of maximum scoring subnetwork

Here  we  use  a  fast  heuristic  approach  to  calculate  an  approximation
to  theoptimal scoring subnetwork.  
An optimal solution can be calculated using the heinz algorithm requiring a 
commercial CPLEX license.

Note: this will run for a couple of minutes so it will be a good idea to cache the
result to not recompute this over and over again.

```{r heinz-bp-gse122425}
module <- BioNet::runFastHeinz(bp.sgr, scores)
module
```

For some reason `BioNet::runFastHeinz` drops the `edgeData` of the input graph,
so we need to re-annotate those to the module.
There doesn't seem to be a convenient way to bulk transfer `edgeData` from one
graph to another, but rather only for one edge and one attribute at a time.

With a little abuse, we help ourselves via accessing slots directly. 
 
```{r}
en <- names(edgeData(module))
edgeDataDefaults(module) <- edgeDataDefaults(bp.sgr)
module@edgeData@data[en] <- edgeData(bp.sgr)[en]
```


## Second reduction step: supervised enrichment analysis of maximum scoring subnetwork 

```{r}
kegg.gs <- EnrichmentBrowser::getGenesets(org = "hsa", db = "kegg", gene.id.type = "UNIPROT")
head(kegg.gs)
go.gs <- EnrichmentBrowser::getGenesets(org = "hsa", db = "go", gene.id.type = "UNIPROT")
head(go.gs)
```

Given the maximum scoring subnetwork and gene sets of interest, we can
now throw all kinds of set-based enrichment analysis (sbea) methods or 
network-based enrichment analysis (nbea) methods onto the problem.

```{r}
EnrichmentBrowser::sbeaMethods()
EnrichmentBrowser::nbeaMethods()
```

Those are conveniently wrapped in the EnrichmentBrowser package and we are
going to try out a couple of them here.

But for now we just take some information from the [paper](https://pubmed.ncbi.nlm.nih.gov/30526041/) to illustrate the concept:

NSUN2 is a nucleolus RNA m5C methyltransferase, regulator of cell proliferation
and cell differentiation, dysfunction associated with intellectual diseases
and cancer; reported to be a downstream target gene of MYC oncogene, a master
regulator of cell proliferation.

Among the pathways showing enrichment in the paper is Ribosome, so as showcase
let's extract nodes participating in the gene set "Ribosome" from the maximum scoring
subnetwork.

```{r}
ind <- grep("Ribosome$", names(kegg.gs))
ribo <- kegg.gs[[ind]]
ribo <- intersect(ribo, graph::nodes(module))
ribo.gr <- graph::subGraph(ribo, module)
ribo.gr
```

Such a gene set induced sub-graph can then further explored in our BioPlex
node data and edge data shiny viewer. 

For that we might be interested in mapping all kind of additional data onto the
nodes and the edges of the network.

For example, we might want to add the Bioplex proteome data onto the graph for
inspection via the shiny graph viewer.

```{r}
bp.prot <- BioPlex::getBioplexProteome()
bp.prot
ribo.gr <- BioPlex::mapSummarizedExperimentOntoGraph(ribo.gr, bp.prot,
    col.names = grep("HEK", colnames(bp.prot), value = TRUE), 
    rowdata.cols = c("nr.peptides", "log2ratio", "adj.pvalue"),
    prefix = "bpprot_")
head(graph::nodeData(ribo.gr), n = 2)
```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```
