---
title: CORUM complex overlap analysis
vignette: >
  % \VignetteIndexEntry{CORUM complex overlap analysis}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r setup, message = FALSE}
library(BioPlex)
library(graph)
library(reticulate)
```

# Data retrieval 

Get the latest version of the 293T PPI network ...

```{r get-bioplex}
bp.293t.3 <- BioPlex::getBioPlex(cell.line = "293T", version = "3.0")
head(bp.293t.3)
```

... and turn into a graph:

```{r bioplex2graph}
gr.293t.3 <- BioPlex::bioplex2graph(bp.293t.3)
```

Obtain the complete set of human protein complexes from CORUM ...

```{r get-corum}
corum.df <- BioPlex::getCorum(set = "core", organism = "Human")
```

... and turn into a `list`:

```{r corum2list}
corum.list <- BioPlex::corum2list(corum.df)
head(corum.list)
```

Identify complexes with at least three subunits:

```{r has-subunit}
hasSubunits <- function(s, gr)
{
    s <- intersect(s, graph::nodes(gr))
    length(s) > 2
}
has.subunits <- vapply(corum.list, hasSubunits, logical(1), gr = gr.293t.3)
table(has.subunits)
```

Identify complexes with at least one subunit targeted as bait:

```{r has-bait}
hasBait <- function(s, df) any(s %in% df$UniprotA)
has.bait <- vapply(corum.list, hasBait, logical(1), df = bp.293t.3)
table(has.bait)
```

Identify complexes with at least one PPI between subunits:

```{r has-edge}
hasEdge <- function(s, gr)
{
    s <- intersect(s, graph::nodes(gr))
    graph::numEdges(graph::subGraph(s, gr)) > 0
}
has.edge <- vapply(corum.list, hasEdge, logical(1), gr = gr.293t.3)
table(has.edge)
```

We then subset the CORUM complexes to those having (i) at least three subunits,
(ii) at least one subunit targeted as bait, and (iii) at least one PPI between subunits.

```{r}
ind <- has.subunits & has.bait & has.edge
table(ind)
sub.corum.list <- corum.list[ind]
```

# Overlap analysis with protein complexes (random sampling)

The [bioplexpy](https://pypi.org/project/bioplexpy/) package implements a 
function for testing overlaps of PPIs with a complex of interest based on
random sampling.

The function samples random subnetworks from the given PPI network, matching the
number of subunits and the bait:prey ratio of the complex being tested.
We then count the number of interactions in each replication, and compare
against the observed number of interactions overlapping with the complex.

Here, we set up a python environment via
[reticulate](https://rstudio.github.io/reticulate/)
that contains the
[bioplexpy](https://pypi.org/project/bioplexpy/) package in order to invoke
the function from within R, facilitating direct exchange of data and results between
the BioPlex R and BioPlex Python packages:

```{r basilisk-start, message = FALSE, warning = FALSE}
reticulate::virtualenv_create("r-reticulate")
reticulate::virtualenv_install("r-reticulate", "bioplexpy")
reticulate::use_virtualenv("r-reticulate")
bp <- reticulate::import("bioplexpy")
```

Here, we use a functionality from the BioPlex Python package to turn the 
BioPlex data into a [NetworkX](https://networkx.org/) graph:

```{r bioplex2graph-py}
gr <- bp$bioplex2graph(bp.293t.3)
gr
```

and also to obtain the subset of CORUM complexes satisfying the above criteria. 

```{r subset-corum}
corum <- bp$getCorum(complex_set = "core", organism = "Human")
sub.corum <- corum[ind,] 
```

Now we carry out the overlap resampling test as implemented in the `bioplexpy`
package for all complexes passing the filter criteria:

```{r perm-test-all}
ps <- vapply(sub.corum$ComplexID,
             function(i) bp$permutation_test_for_CORUM_complex(gr, sub.corum, i, 100),
             numeric(1))
head(ps)
summary(ps)
```

We can now take a peak at the p-value distribution from the overlap permutation
test, observing a strong concentration near zero, confirming that, as expected,
many complexes are enriched for PPIs - when compared to random subsets of the
overall PPIs network matching the complexes in size and composition.

```{r p-hist}
hist(ps, breaks = 20, xlab = "p-value", ylab = "frequency", col = "#00AFBB")
```

# Overlap analysis with protein complexes (network randomization)

The resampling test implemented in the BioPlex Python package addresses the question:
given a specific network topology, how likely is it that *n* nodes are connected
by at least *m* edges? In the spirit of a competitive enrichment test, we
therefore assess whether the connectivity within the protein of interest exceeds
the connectivity outside of the the protein set of interest (= overall / background
connectivity). 

It is also possible to test each complex for enrichment of PPIs based on
network randomization as implemented in the
[BiRewire](https://bioconductor.org/packages/BiRewire)
 package. This addresses the question: in networks of different topology but the
same node degree distribution, how likely is it that *n* nodes are connected by
at least *m* edges? The network randomization test thus randomizes the overall
network while keeping the protein set of interest fixed (in contrast to the resampling
test, which randomizes the protein set of interest while keeping the overall network fixed).

This approach is incorporated in the `testConnectivity` function, which we apply
here to test all CORUM complexes that were passing the filter criteria above. 
We therefore randomize the PPI network a defined number of times (here: 100 times),
and calculate for each complex how often the number of edges in a randomized
configuration exceeded the number of edges observed for the true PPI network.  

```{r net-rand-test, message = FALSE}
ps2 <- BioPlexAnalysis::testConnectivity(sub.corum.list, 
                                         gr.293t.3,
                                         nr.reps = 10)
head(ps2)
summary(ps2)
```

As before we can take a peak at the *p*-value distribution for the network
randomization test, which displays as for the resampling test a strong enrichment
of PPIs across almost all CORUM complexes. 

```{r p-hist2}
hist(ps2, breaks = 20, xlab = "p-value", ylab = "frequency", col = "#00AFBB")
```


