---
title: Using a GraphFrames backend
vignette: >
  % \VignetteIndexEntry{Using a GraphFrames backend}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r, message = FALSE}
library(BioPlex)
library(BioPlexAnalysis)
library(sparklyr)
library(graphframes)
library(dplyr)
```

# Connecting to Spark 

For working GraphFrames, we need a connection to Spark.
For computation on a computer cluster, this can be achieved by connecting to
the spark installation on the cluster.

However, for demonstration purposes we are using here a local connection,
that will require a local Spark installation.

This can be achieved via (need to be carried out only once):

```{r spark-install, eval = FALSE}
sparklyr::spark_install("3.0")
```

We can connect to both local instances of Spark as well as remote Spark clusters.
Here we connect to our local installation of Spark via

```{r spark-connect, warning = FALSE}
sc <- sparklyr::spark_connect(master = "local", version = "3.0")
```

# Store BioPlex PPIs in a GraphFrame

Get the latest version of the 293T PPI network:

```{r}
bp.293t <- BioPlex::getBioPlex(cell.line = "293T", version = "3.0")
head(bp.293t)
```

and turn into a graph object:

```{r}
bp.gr <- BioPlex::bioplex2graph(bp.293t)
bp.gr
```

Switch to a graphframes backend:

```{r, warning = FALSE}
gf <- BioPlexAnalysis::graph2graphframe(bp.gr, sc)
gf
```

# PageRank

[PageRank](https://en.wikipedia.org/wiki/PageRank) is an algorithm used in
Google Search for ranking websites in their results, but it has been adopted
also for other purposes. According to Google, PageRank works by counting the number
and quality of links to a page to determine a rough estimate of how important
the website is. 
The underlying assumption is that more important websites are likely to receive
more links from other websites.

For the analysis if PPI networks, *personalized* PageRank seems to be capable to
robustly evaluate the importance of the vertices of a network, relatively to some
already known relevant nodes
([Ivan and Grolmusz, 2011](https://academic.oup.com/bioinformatics/article/27/3/405/321946)).

As an example we look at the ACADVL gene. Mutations in ACADVL are associated with
very long-chain acyl-coenzyme A dehydrogenase deficiency.

We first look up the corresponding node ID:

```{r}
dplyr::filter(graphframes::gf_vertices(gf), symbol == "ACADVL")
```

And then apply the personalized PageRank as implemented in the `graphframes`
package.

```{r, warning = FALSE}
gf <- graphframes::gf_pagerank(gf, 
                               reset_prob = 0.15,
                               max_iter = 10L,
                               source_id = "P49748")
gf
```

Inspect the results:

```{r}
dplyr::filter(graphframes::gf_vertices(gf), pagerank > 0)
dplyr::filter(graphframes::gf_edges(gf), src == "P49748")
```

Switch back to a `graphNEL` backend:

```{r, warning = FALSE}
gr <- BioPlexAnalysis::graphframe2graph(gf)
gr
head(graph::nodeData(gr), n = 2)
head(graph::edgeData(gr), n = 2)
```
