---
vignette: >
  % \VignetteIndexEntry{Using a GraphFrames backend}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r, message = FALSE}
library(BioPlex)
library(BioPlexAnalysis)
library(sparklyr)
library(graphframes)
```

# Connecting to Spark 

For working GraphFrames, we need a connection to Spark.
For computation on a computer cluster, this can be achieved by connecting to
the spark installation on the cluster.

However, for demonstration purposes we are using here a local connection,
that will require a local Spark installation.

This can be achieved via (need to be carried out only once):

```{r spark-install, eval = FALSE}
sparklyr::spark_install("3.0")
```

We can connect to both local instances of Spark as well as remote Spark clusters.
Here we connect to our local installation of Spark via

```{r}
sc <- sparklyr::spark_connect(master = "local", version = "3.0")
```

# Store BioPlex PPIs in a GraphFrame

Get the latest version of the 293T PPI network:

```{r}
bp.293t <- BioPlex::getBioPlex(cell.line = "293T", version = "3.0")
head(bp.293t)
```

and turn into a graph object:

```{r}
bp.gr <- BioPlex::bioplex2graph(bp.293t)
bp.gr
```

Switch to a graphframes backend:

```{r}
gf <- bioplex2graphframe(bp.gr, sc)
gf
```

