---
title: Mapping PPIs onto 3D protein structures
vignette: >
  % \VignetteIndexEntry{Mapping PPIs onto 3D protein structures}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

The [bioplexpy](https://pypi.org/project/bioplexpy/) package implements functions
to calculate and visualize physical interactions between chains of a PDB structure
and compare with BioPlex PPIs detected in the AP-MS data.

This can be used to identify subunits of a protein complex that are close enough
to physically interact. This analysis (1) maps a set of UniProt IDs, corresponding
to the subunits of the complex, to a PDB structure, and (2) infers physical interactions
within a structure by calculating the distance between the atoms of different
chains within that structure.  

Here, we demonstrate how to invoke this functionality from within R, and how to
incorporate results into subsequent analysis and visualization in R. 

We start by loading the required packages.

```{r, message = FALSE}
library(BioPlex)
library(graph)
library(bio3d)
library(reticulate)
library(r3dmol)
```

# Mapping from a set of UniProt IDs to a PDB structure

We use data from the [SIFTS](https://www.ebi.ac.uk/pdbe/docs/sifts/quick.html)
project to map from CORUM complex subunits given as UniProt IDs 
to PDB structures:

```{r}
url <- "ftp://ftp.ebi.ac.uk/pub/databases/msd/sifts/flatfiles/csv/uniprot_pdb.csv.gz"
dest <- basename(url)
download.file(url, destfile = dest)
df <- read.csv("uniprot_pdb.csv.gz", skip = 1)
file.remove(dest)
head(df)
```

Turn into a mapping:

```{r}
unip2pdb <- strsplit(df$PDB, ";")
names(unip2pdb) <- df$SP_PRIMARY
head(unip2pdb)
```

Having illustrated the basic principle underlying the mapping from a CORUM complex
to a PDB structure, we now set up a python environment via
[reticulate](https://rstudio.github.io/reticulate/)
that contains the
[bioplexpy](https://pypi.org/project/bioplexpy/) package in order to invoke
the functionality from within R, facilitating direct exchange of data and results between
the BioPlex R and BioPlex Python packages:

```{r basilisk-start, message = FALSE, warning = FALSE}
reticulate::virtualenv_create("r-reticulate")
reticulate::virtualenv_install("r-reticulate", "bioplexpy")
reticulate::use_virtualenv("r-reticulate")
bp <- reticulate::import("bioplexpy")
```

In the following, we focus on the TFIIH core complex (PDB ID: [6nmi](https://www.rcsb.org/structure/6NMI)).

We first get the set of UniProt IDs corresponding to this CORUM complex ID.

```{r uniprots-from-corum}
corum <- bp$getCorum(complex_set = "core", organism = "Human")
unips <- bp$get_UniProts_from_CORUM(corum, 107) 
unips
```

and then also get the set of PDB IDs corresponding to the obtained set of UniProt IDs
based on the SIFTS mapping. 

```{r up2pdb}
pdbs <- bp$get_PDB_from_UniProts(unips)
head(pdbs)
```

As there is often no 1:1 mapping from a set of UniProt IDs to a certain PDB structure,
it is instructive to inspect the title of the PDB entry and use prior knowledge
about the PDB structure for a complex of interest where available.

Here, we accordingly chose [6nmi](https://www.rcsb.org/structure/6NMI) as the PDB ID
representing the TFIIH core complex.

We can use functionality from the 
[bio3d](http://thegrantlab.org/bio3d/) package to get the PDB structure for the
TFIIH core complex (PDB ID: [6nmi](https://www.rcsb.org/structure/6NMI)).

```{r}
pdb.id <- "6nmi"
pdb.file <- bio3d::get.pdb(pdb.id)
pdb <- bio3d::read.pdb(pdb.file)
pdb
str(pdb)
```

Provide a different color for each chain:

```{r}
chains <- unlist(pdb$remark$biomat$chain)
chains
nr.chains <- length(chains)
chain.colors <- RColorBrewer::brewer.pal(8, "Set1")
chain.colors <- tolower(chain.colors)
```

And visualize the structure using functionality from the 
[r3dmol](https://cran.r-project.org/web/packages/r3dmol/index.html) package:

```{r}
# Set up the initial viewer
viewer <- r3dmol(
  id = "",
  elementId = "demo"
) %>%
  # Add model to scene
  m_add_model(data = m_bio3d(pdb), format = "pdb") %>%
  # Zoom to encompass the whole scene
  m_zoom_to() %>%
  # Set style of structures
  m_set_style(style = m_style_cartoon(color = "#00cc96"))
for(i in seq_len(nr.chains))
  viewer <- m_set_style(viewer,
    sel = m_sel(chain = LETTERS[i]),
    style = m_style_cartoon(color = chain.colors[i])
  ) 
viewer %>% m_rotate(angle = 90, axis = "y") %>% m_spin()
```

# Identify interacting chains within a protein structure

We infer physical interactions within a structure by calculating the distance
between the atoms of different chains within that structure. 
The coordinates of each atom can be obtained directly from the PDB file.

```{r atoms}
rel.cols <- c("type", "chain", "x", "y", "z")
pos <- pdb$atom[,rel.cols]
head(pos)
```

If the minimum distance between atoms of two chains is below a given threshold
(default: 6 angstrom), these proteins are defined to interact directly; all other
pairs of proteins that occurred in the same structure are assumed to interact indirectly.

```{r ia-chains}
ia.chains <- bp$get_interacting_chains_from_PDB('6NMI', '.', 6)
ia.chains
```

For example, this demonstrates that chain A (red) and chain B (blue) are in close
proximity, and that both chains are also close to chain H (pink), as evident from
the structure view above. 

Once we have obtained the list of interacting chains, we map each chain to a
UNIPROT ID, again based on data from the SIFTS project.

```{r chain2unip}
chain2unip <- bp$list_uniprot_pdb_mappings('6NMI')
chain2unip
```

This mapping now allows us to produce a list of interacting chains using
their corresponding UNIPROT ID.

```{r ia-unips}
chain2unip <- unlist(chain2unip)
ia.unips <- lapply(ia.chains, function(x) unname(chain2unip[x]))
ia.unips
```

We reshape this `list` to a `data.frame`:

```{r ia-df}
ia1 <- vapply(ia.unips, `[`, character(1), x = 1)
ia2 <- vapply(ia.unips, `[`, character(1), x = 2)
ia.df <- data.frame(CHAIN1 = ia1, CHAIN2 = ia2)
ia.df
``` 

# Identify interactions that are also detected in the BioPlex network

Having inferred physical interactions between subunits of the TFIIH core complex,
we can now investigate which of these interactions have been detected in the BioPlex
PPI data.

We therefore obtain the latest version of the BioPlex network for the 293T cell line.

```{r bp-293t}
bp.293t <- BioPlex::getBioPlex(cell.line = "293T", version = "3.0")
```

Now we restrict the inferred physical interactions between subunits of the TFIIH core complex
to those also found in the BioPlex 293T network.

```{r}
ia.comb1 <- paste(ia.df[,1], ia.df[,2], sep = "-")
ia.comb2 <- paste(ia.df[,2], ia.df[,1], sep = "-")
bp.comb <- paste(bp.293t$UniprotA, bp.293t$UniprotB, sep = "-")
ind1 <- ia.comb1 %in% bp.comb
ind2 <- ia.comb2 %in% bp.comb
ind <- ind1 | ind2
ia.df[ind,]
```
