---
vignette: >
  % \VignetteIndexEntry{PFAM domain-domain association analysis}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    mathjax: null   
---

# Setup

```{r, message = FALSE}
library(BioPlex)
library(AnnotationDbi)
library(AnnotationHub)
library(graph)
```

Get the latest version of the 293T PPI network:

```{r}
bp.293t <- BioPlex::getBioPlex(cell.line = "293T", version = "3.0")
```

and turn into a graph object:

```{r}
bp.gr <- BioPlex::bioplex2graph(bp.293t)
```

Annotate PFAM domains to the node metadata:

Connect to 
[AnnotationHub](http://bioconductor.org/packages/AnnotationHub):

```{r ahub, message = FALSE}
ah <- AnnotationHub::AnnotationHub()
```

OrgDb package for human:
```{r orgdb, message = FALSE}
AnnotationHub::query(ah, c("orgDb", "Homo sapiens"))
orgdb <- ah[["AH92581"]]
orgdb
AnnotationDbi::keytypes(orgdb)
```

```{r pfam}
bp.gr <- BioPlex::annotatePFAM(bp.gr, orgdb)
```

Domain-domain association analysis:

1. create a map from PFAM to UNIPROT

```{r}
unip2pfam <- graph::nodeData(bp.gr, graph::nodes(bp.gr), "PFAM")
pfam2unip <- stack(unip2pfam)
pfam2unip <- split(as.character(pfam2unip$ind), pfam2unip$values)
head(pfam2unip, 2)
```

2. create a map from UNIPROT to PFAMs of interaction partners

```{r}
unip2iapfams <- lapply(graph::nodes(bp.gr), function(n) graph::nodeData(bp.gr, graph::edges(bp.gr)[[n]], "PFAM"))
unip2iapfams <- lapply(unip2iapfams, unlist)
names(unip2iapfams) <- graph::nodes(bp.gr)
```

3. for each pfam, get the pfams of interacting proteins

```{r}
pfam2iapfams <- lapply(pfam2unip, function(ps) unlist(unip2iapfams[ps]))
pfam2iapfams <- pfam2iapfams[lengths(pfam2iapfams) > 0]
pfam2pfam <- stack(pfam2iapfams)
pfam2pfam$ind <- as.character(pfam2pfam$ind) 
pfam2pfam <- pfam2pfam[!is.na(pfam2pfam$values),]
pfam2pfam <- as.matrix(pfam2pfam)
head(pfam2pfam)
```

4. compute an all-against-all matrix storing pfam2pfam interaction counts:

4a. initialize + turn pfam2pfam mapping into an integer index mapping for efficiency 
```{r} 
pfams <- unique(as.vector(pfam2pfam))
len <- length(pfams)
ia.mat <- matrix(0, nrow = len, ncol = len)
rownames(ia.mat) <- colnames(ia.mat) <- pfams
ind1 <- match(pfam2pfam[,1], pfams)
ind2 <- match(pfam2pfam[,2], pfams)
pfam2pfam <- cbind(ind1, ind2)
head(pfam2pfam)
```

4b. loop over all pfam-pfam pair and count the PPIs connecting them:
```{r}
for(i in seq_len(len))
{
    if(i %% 100 == 0) message(i)
    p2p <- pfam2pfam[pfam2pfam[,1] == i |  pfam2pfam[,2] == i,,drop = FALSE]
    if(nrow(p2p))
    {  
      for(j in i:len)
      {
          if(i == j) ia.mat[i,j] <- sum(p2p[,1] == j & p2p[,2] == j)
          else ia.mat[i,j] <- sum(p2p[,1] == j | p2p[,2] == j)
      }
    }  
}
```

Inspect the result:

```{r}
dim(ia.mat)
ia.mat[1:5,1:5]
```

5. filter out pfams that have not at least one pfam with 2 interactions 

```{r}
gr1 <- apply(ia.mat, 1, function(x) any(x > 1))
ia.mat <- ia.mat[gr1 , gr1] 
dim(ia.mat)
```

6. calculate 2x2 contigency tables for all domain pairs:

6a. Let's do some precomputations of the margins for efficiency
```{r} 
total <- sum(as.vector(ia.mat))
m <- ia.mat
m[lower.tri(m)] <- t(m)[lower.tri(m)]
rs <- rowSums(m)
```

6b. let's define a function that computes for each domain pair:
    1) the number of interactions connecting both domains;
    2) the numbers of interactions involving either domain individually; and
    3) the number of interactions not involving either domain

```{r}
getContingency <- function(p1, p2, m, rs, total) 
{
    both <- m[p1,p2]
    one <- rs[p1] - both
    two <- rs[p2] - both
    all <- total - one - two
    unname(c(both, one, two, all))
}
getContingency("PF02874", "PF00006", m, rs, total)
```

6c. now let's get all possible domain pairs and restrict to those pairs connected
by at least 2 PPI's:

```{r}
combs <- combn(rownames(m), 2)
ind <- apply(combs, 2, function(x) m[x[1], x[2]] > 1)
combs <- combs[,ind]
dim(combs)
```

6d. calculate contigency for all remaining domain pairs:

```{r}
conts <- apply(combs, 2, function(x) getContingency(x[1], x[2], m, rs, total)) 
conts <- t(conts)
head(conts)
```

7. Now we test each domain pair for over-representation of connecting PPI's
based on the hypergeometric distribution.
This corresponds to a one-sided version of Fisher’s exact test:

```{r}
fisherp <- function(x) fisher.test(matrix(x, nrow = 2), 
                       alternative = "greater")$p.value 
fisherp(conts[1,])
ps <- apply(conts, 1, fisherp)
head(ps)
hist(ps, breaks = 50)
```

8. Multiple testing correction:
```{r}
adjp <- p.adjust(ps, method = "BH")
```

RESULT:
```{r}
res <- data.frame(PFAM1 = combs[1,], 
                  PFAM2 = combs[2,],
                  PVAL = ps,
                  ADJ.PVAL = adjp)
res <- res[order(ps),]
head(res)
```

9. Get human readable description for the PFAM domain IDs:

```{r}
library(PFAM.db)
id2de <- as.list(PFAMDE)
```

What are the top 10 interacting domains:

```{r}
cbind(id2de[head(res$PFAM1, 10)],
      id2de[head(res$PFAM2, 10)])
```

Can we do some of this vectorized ?

```{r}
.f <- function(x, rs, total)
{
  one <- rs[1] - x
  two <- rs[2] - x
  all <- total - one - two
  cbind(x, one, two, all)
}      
```


